manuscript

### 2023-05-25 manuscript.rmd
added text and references

---
title: "manuscript"
format: html
editor: visual
---

@@

\[Introduce the importance of passive acoustic monitoring on this type of study and gaps in visual survey methods\]\
Passive acoustic monitoring is a tool that allows us to record and listen to the soundscape. Beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen. One type of passive acoustic recorder called a drifting buoy allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters; it can drift freely with the currents for days or weeks at a time. The goal with passive acoustic monitoring is to improve current species distribution models of beaked and sperm whales that utilize visual survey methods only. Its basic components consist of a satellite transmitter above water to monitor GPS information and an acoustic recorder attached by line at \~100m of depth. In fact, most of what is known about these deep divers is because of acoustic monitoring.

@@

\[Environmental Data\]\
Environmetal and oceanographic variables thought to be predictors of BW/SW habitat were extracted from NOAA's Environmental Research Division Data Access Program (ERDDAP) and Hybrid Coordinate Ocean Model (HYCOM) using matchEnvData function in package PAMmisc. Variables at depth were calculated using function from Taiki and formula from McCullough et al 2021. Distance to slope and bathymetric slope were calculated from a bathymetry TIF taken from GEBCO. Can show table of variables and justification by papers (Virgili et al. 2022).

@@

[Example of GAM]{.underline}: Manocci et al. 2015.

@@

>>>>

---
title: "manuscript"
format: html
editor: visual
bibliography: references.bib
---

@@

\[Introduce the importance of passive acoustic monitoring on this type of study and gaps in visual survey methods\]\
Passive acoustic monitoring is a tool that allows us to record and listen to the soundscape. Beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen. One type of passive acoustic recorder called a drifting buoy allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters. Studying beaked whales can be challenging due to how offshore they live (@robbins2022). It can drift freely with the currents for days or weeks at a time. The goal with passive acoustic monitoring is to improve current species distribution models of beaked and sperm whales that utilize visual survey methods only. Historically, it has been challenging to record foraging calls at several hundred meters of depth (\@barlow).

Its basic components consist of a satellite transmitter above water to monitor GPS information and an acoustic recorder attached by line at \~100m of depth. In fact, most of what is known about these deep divers is because of acoustic monitoring.

@@

\[Environmental Data\]\
Environmetal and oceanographic variables thought to be predictors of BW/SW habitat were extracted from NOAA's Environmental Research Division Data Access Program (ERDDAP) and Hybrid Coordinate Ocean Model (HYCOM) using matchEnvData function in package PAMmisc. Variables at depth were calculated using function from Taiki and formula from McCullough et al 2021. Distance to slope and bathymetric slope were calculated from a bathymetry TIF taken from GEBCO.

\[Can show table of variables and justification by papers (Virgili et al. 2022).\]

| Covariate | Unit | Source                                                   | Justification |
|-----------|------|----------------------------------------------------------|---------------|
|           |      | \[can use A, B, C to avoid repetition, cite in caption\] |               |
|           |      |                                                          |               |
|           |      |                                                          |               |
|           |      |                                                          |               |
|           |      |                                                          |               |
|           |      |                                                          |               |
|           |      |                                                          |               |
|           |      |                                                          |               |
|           |      |                                                          |               |
|           |      |                                                          |               |
|           |      |                                                          |               |
|           |      |                                                          |               |
|           |      |                                                          |               |
|           |      |                                                          |               |

@@

[Example of GAM]{.underline}: Manocci et al. 2015. @mannocci2015

AIC explanation @symonds2011

@@

Are BW or SW known to have seasonal movements? @robbins2022 suggest they do, though there's conflicting evidence.
